A Bayesian search to find \\high-mass black holes in LIGO data

### 2020-08-19 main.tex
add introduction

@@

\title{A Bayesian search to find \\high-mass black holes using LIGO data}% Force line breaks with \\

@@

\begin{abstract}
The detection of high mass black holes ( $>100$ M${}_\odot$) will shed light on the formation of supermassive black holes and thus galaxy formation. Although LIGO is sensitive to the merger of binary black holes with total masses up to 500 M${}_\odot$, the largest total mass detected so far is approximately 80 M${}_\odot$. A possible explanation for the absence of high mass events may be due to their misclassification as short-duration instrumental noise transients. Short-duration instrumental transients mimic the short-duration gravitational-wave signals from high-mass binary black hole mergers. Here we demonstrate that a new search method utilising Bayesian inference could be a more sensitive tool for detecting high-mass binary black hole mergers as compared to traditional match-filtering. We have applied this technique on the high-mass triggers during LIGO's second observing run to investigate the possibility of discovering new gravitational-wave signals from an entirely new class of high mass black hole binaries.

@@

Since the 1970s, there has been an accumulation of evidence for stellar-mass and supermassive black holes. In 2019, the Event Horizon Telescope provided the first visual evidence of the supermassive black hole M87  \cite{m87photo}.  As of January 2020, LIGO has confirmed ten binary black hole systems, and more than fifty public binary black hole alerts \cite{gw190814}. These various discoveries have firmly established the existence of stellar-mass black holes, supermassive black holes and binary black hole systems.  Interestingly, there is still no direct evidence for intermediate-mass black holes, the black holes that lie in between stellar-mass and supermassive black hole systems with masses between $10^2-10^6 \text{M}_\odot$. Thus, the existence of intermediate-mass black holes is still speculative. \\

If intermediate-mass black holes are present, the gravitational waves emitted from the merger of a binary black hole system with at least one intermediate-mass black hole (up to a mass of $400 \text{M}_\odot$) will be detectable by the ground-based gravitational-wave network. According to \citet{fregeau2006imbhbRatePrediction, mandel2008rates,rodriguez2015bbhRatePredictions}, gravitational waves from such systems should occur at a rate of 0-10 yr\(^{-1}\) .  However, even after \citet{salemi2019search}'s targeted match-filter based search for gravitational waves from high-mass black holes the largest total mass detected so far is approximately 80 M\({}_\odot\) \cite{abbott2019gwtc}. A possible explanation for the absence of high mass events may be
due to their misclassification as short-duration instrumental noise transients \cite{blip_glitches}.

% PyCBC asks for the 5 citations:
% http://pycbc.org/pycbc/latest/html/credit.html#searches-for-compact-binary-coalescence

Search algorithms for gravitational waves from coalescing compact binaries, such as PyCBC \cite{pycbc_code}, have tried to account for these short-duration glitches \cite{pycbc_short_duration_transients}. At its core, PyCBC performs a matched-filter search for binary merger signals using a bank of gravitational-wave template waveforms \cite{pycbc_og0, pycbc_og1, pycbc_og2, pycbc_og3, pycbc_og4, pycbc_og5}.  To account for the short-duration glitches,

>>>>

\newcommand{\prior}{{\sc prior}\xspace}
\newcommand{\bilby}{{\sc Bilby}\xspace}
\newcommand{\bilbypipe}{{\sc bilby\_pipe}\xspace}
\newcommand{\bilbypipegdb}{\texttt{bilby\_pipe\_gracedb}\xspace}
\newcommand{\pbilby}{{\sc pBilby}\xspace}
\newcommand{\dynesty}{{\sc dynesty}\xspace}
\newcommand{\cpnest}{{\sc cpnest}\xspace}
\newcommand{\ptemcee}{{\sc ptemcee}\xspace}
\newcommand{\gwpy}{{\sc GWpy}\xspace}
\newcommand{\imrphenomp}{{\sc IMRPhenomPv2}\xspace}
\newcommand{\pycbc}{{\sc PyCBC}\xspace}
\newcommand{\msun}{{M${}_\odot$}\xspace}

@@

\title{A Bayesian search to find \\high-mass black holes in  LIGO data}% Force line breaks with \\

@@

\begin{abstract}
The detection of high mass black holes ( $>100$ \msun) will shed light on the formation of supermassive black holes and thus galaxy formation. Although LIGO is sensitive to the merger of binary black holes with total masses up to 500 \msun, the largest total mass detected so far is approximately 80 \msun. A possible explanation for the absence of high mass events may be due to their misclassification as short-duration instrumental noise transients. Short-duration instrumental transients mimic the short-duration gravitational-wave signals from high-mass binary black hole mergers. Here we demonstrate that a new search method utilising Bayesian inference could be a more sensitive tool for detecting high-mass binary black hole mergers as compared to traditional match-filtering. We have applied this technique on the high-mass triggers during LIGO's second observing run to investigate the possibility of discovering new gravitational-wave signals from an entirely new class of high mass black hole binaries.

@@

Since the 1970s, there has been an accumulation of evidence for stellar-mass and supermassive black holes. In 2019, the Event Horizon Telescope provided the first visual evidence of the supermassive black hole M87  \cite{m87photo}.  As of January 2020, LIGO, Virgo and IAS have confirmed more than a dozen binary black hole systems \cite{GWTC1, IAS0, IAS1, IAS2}. These various discoveries have firmly established the existence of stellar-mass black holes, supermassive black holes and binary black hole systems.  Interestingly, there is still no direct evidence for intermediate-mass black holes, the black holes that lie in between stellar-mass and supermassive black hole systems with masses between $10^2-10^6$ \msun. Thus, the existence of intermediate-mass black holes is still speculative. \\

If intermediate-mass black holes are present, the gravitational waves emitted from the merger of a binary black hole system with at least one intermediate-mass black hole (up to a mass of $400$ \msun) will be detectable by the ground-based gravitational-wave network. Gravitational waves from such systems should occur at a rate of $0-10\text {yr}^{-1}$ \cite{fregeau2006imbhbRatePrediction, mandel2008rates,rodriguez2015bbhRatePredictions}.  However, even after conducting a targeted match-filter based search for gravitational waves from intermediate-mass black holes the largest total mass detected so far is approximately $80$ \msun \cite{salemi2019search, abbott2019gwtc}. A possible explanation for the absence of intermediate-mass events may be due to their misclassification as short-duration instrumental noise transients \cite{blip_glitches}. These glitches can mimic astrophysical signals and hence decrease the significance of true gravitational wave events.

One method to account for glitches while searching for gravitational waves from coalescing compact binaries is by utilising a signal consistency test \cite{pycbc_short_duration_transients} or a signal coherence test amongst a network of detectors \cite{bci, bcr_paper,  bcr_gw151216, bayesian_odds}. A signal-coherence test such as the Bayesian coherence ratio (BCR) \cite{bcr_paper} provides an odds ratio comparing the probability that the data contains an astrophysical signal versus the probability that the data contains an incoherent glitch or noise. Although an odds ratio, the BCR can be used  as a ranking statistic to calculate the statistical significance of candidate gravitational-wave signals from compact binary coalescences in the detector data \cite{bcr_paper}.

In this paper we utilise the BCR to calculate $p_\text{astro}$, the probability that a signal is of astrophysical origin [cite],  for all the high-mass (systems with total masses in the range of $55-500$ \msun) candidate gravitational-wave signals from compact binary coalescences in the detector data recorded during aLIGO's second observing run (O2).  We find that (a) our search does not identify any new stellar mass or intermediate mass black holes; (b) all events reported in the GWTC-1, including GW170729 (an event with disputed $p_\text{astro}$ amongst various search pipelines) have high $p_\text{astro}$; and that (c) events

is an odds ratio

presents the method and results from analysing the candidate events with

\section{PyCBC}
% PyCBC asks for the 5 citations:
% http://pycbc.org/pycbc/latest/html/credit.html#searches-for-compact-binary-coalescence

 such as \pycbc \cite{pycbc_code} have tried to account for these short-duration glitches \cite{pycbc_short_duration_transients}. 
Search algorithms for gravitational waves from coalescing compact binaries, such as PyCBC \cite{pycbc_code}, have tried to account for these short-duration glitches \cite{pycbc_short_duration_transients}. At its core, PyCBC performs a matched-filter search for binary merger signals using a bank of gravitational-wave template waveforms \cite{pycbc_og0, pycbc_og1, pycbc_og2, pycbc_og3, pycbc_og4, pycbc_og5}.  To account for the short-duration glitches,
